### 2016-08-25 robust-software.tex
Re-wrote introducation and added references

\begin{quote}
\begin{center}\textbf{Abstract}\end{center}
Software produced for research,
published and otherwise,
suffers from a number of common problems
that make it difficult or impossible to run outside the original institution,
or even off the primary developer's computer.
We present ten simple rules to make such software robust enough to run anywhere
and thereby delight your users and collaborators.
\end{quote}

@@

As a relatively young field, bioinformatics is full of newly developed software.
\fixme{MT}{examples here}
Efforts such as
\withurl{the ELIXIR tools and data registry}{http://dx.doi.org/10.1093/nar/gkv1116}
and the
\withurl{Bioinformatics Links Directory}{https://bioinformatics.ca/links\_directory/}
\cite{brazas2012}
have made efforts towards cataloguing it: as of May 2016, the former
has 2500 entries and the latter 1700. Those numbers constantly
grow as both trainees and experienced practitioners produce new
software to support their research.

Typically, this software is developed by one person and 
produces excellent results in their hands. But what happens after the
student leaves someone else from their group wants to use the program? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
Its new user is often faced with two unpalatable options:
hack the existing code to make it work, or start over.

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. In particular,
robust software:

@@

These rules need not be applied to \emph{every} coding effort:
as the saying goes, not everything worth doing is worth doing right (or right away).
Code  that is used once to answer a specific question related to a specific dataset
doesn't require comprehensive documentation,
and the only sensible way to test it may well be to run it on the dataset in question.
However, if a script is dusted off and run three or four
times for slightly different purposes,
crucial to a publication,
or being passed on to someone else,
it may be time to apply ten simple rules for robust software.

These simple rules are also necessary steps toward creating a reusable
library that can be shared through a site like CPAN or CRAN.  They are
generic and can be applied to all languages, libraries, packages,
documentation styles, and operating systems for both closed-source and
open-source sofwtare.  Whether the aim is as simple as sharing the
code with collaborators or as complex as using the software in a
production analysis environment, increasing the robustness of your
software decreases headaches all around.

>>>>

\begin{quote}
\begin{center}\textbf{Abstract}\end{center}
Software produced for research,
published and otherwise,
suffers from a number of common problems
that make it difficult or impossible to run outside the original institution,
or even off the primary developer's computer.
We present ten simple rules to make such software robust enough to run anywhere,
and inspire confidence in your reproducibility,
and thereby delight your users and collaborators.
\end{quote}

@@

Typically, scientific software is developed and used by a single person,
usually a graduate student or postdoc, and produces excellent results in their
hands\cite{prins2015}. 
But what happens
when someone else wants to use their software? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
The new user is then faced with two unpalatable options:
hack the existing code to make it work, or start over. Being unable to
replicate results is so common that one publication refers to it as "a rite of
passage"\cite{baker2016}.

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. With a lack of
robustness also comes a lack of reproducibility and many duplicated efforts,
slowing the pace of research\cite{prabhu2011,lawlor2015}.
Bioinformatics is full of such efforts, catalogued by efforts such as
the ELIXIR tools and data registry\cite{ison2016}
and the Bioinformatics Links Directory \cite{brazas2012}. In the Bioinformatics
Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.

This problem is not unique to bioinformatics, or even to computing\cite{baker2016}. 
Best practices in software engineering specifically aim to increase software
robustness. However, most bioinformaticians come from related disciplines and
are trained on-the-job or otherwise informally\cite{prins2015,atwood2015}. Even
the existing programs and initiatives rarely have the time to cover engineering
topics in-depth, especially since the field is so broad and rapidly
developing\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
and knowledge are taught, developing robust software is not directly rewarded
in science and funding is difficult to come by\cite{prins2015}. Some proposed
solutions to this problem, in addition to restructuring educational programs,
include hiring dedicated software engineers \cite{lawlor2015,sanders2008},
partnering with private sector or grassroots organizations\cite{prins2015,ison2016},
or using specific technical tools like containerization or cloud
computing\cite{afgan2016,howe2012}. Each of these requires time and, in some
cases, institutional change.

You don't need to be a full-time programmer to write robust software. In fact,
some of the best, most reliable pieces of software in many scientific
communities are written by researchers \cite{prabhu2011,sanders2008}. 
The most widely-used software packages tend to adopt strong software
engineering approaches, have high standards of reproducibility, good testing
practices, and inter-institutional collaborations that grow into strong user
bases. In the bioinformatics community, Bioconductor and Galaxy follow this
path\cite{Gentleman2004,afgan2016}.

That said, not \emph{every} coding effort requires such rigor:
as the saying goes, not everything worth doing is worth doing right (or right away).
Code that is used once to answer a specific question related to a specific dataset
doesn't require comprehensive documentation or flexible configuration,
and the only sensible way to test it may well be to run it on the dataset in
question. Exploratory analysis is an iterative process that is developed
quick and revised often\cite{lawlor2015,sanders2008}. 
However, if a script is dusted off and run three or four
times for slightly different purposes,
is crucial to a publication or a lab,
or being passed on to someone else,
it may be time to make your software more robust.
In particular, robust software:

@@

These simple rules are also necessary steps toward creating a reusable
library that can be shared and reused through a site like CPAN or CRAN.  They are
generic and can be applied to all languages, libraries, packages,
documentation styles, and operating systems for both closed-source and
open-source software.  
Whether the aim is as simple as sharing the
code with collaborators or as complex as using the software in a
production analysis environment, increasing the robustness of your
software decreases headaches all around.
